% !TEX root = paper.tex
\section{Related Work}\label{sec:related}

\subsection{%
	Diffusion Models for Image and Video Synthesis}\label{sub:related:diffusion_models}

Diffusion models
have emerged
as the dominant paradigm
for high-fidelity image generation
and increasingly constitute the foundation for video synthesis applications.
Foundational works establish generation
as iterative denoising\cite{%
	pmlr-v37-sohl-dickstein15, ho2020denoising},
providing strong baselines
for image quality benchmarks.
Score-based formulations
further connect diffusion
with denoising score matching
and establish continuous-time dynamics
\cite{%
	song2019generative},
while recent architectural innovations
push diffusion beyond GAN performance
on standard image benchmarks\cite{dhariwal2021diffusionmodelsbeatgans}.
For video synthesis,
diffusion models have been extended
to spatiotemporal settings,
achieving strong visual quality
through scalable attention mechanisms
and efficient broadcasting strategies
\cite{zhao2025realtimevideogenerationpyramid}.

\subsection{%
	Acceleration Methods for Diffusion Models}\label{sub:related:acceleration_methods}

Despite the remarkable generation quality
achieved by diffusion models
in image and video synthesis,
their inherently sequential inference procedure
and substantial computational requirements
pose significant barriers
to practical deployment
and real-time applications.
Existing approaches
to accelerating diffusion model inference
can be broadly categorized as follows:

\textbf{Sampling Acceleration.}
One fundamental approach to acceleration
reduces the number of denoising steps
via fast sampling algorithms,
including DDIM \cite{song2020denoising},
ODE/SDE solvers \cite{Karras2022edm,Lu_2025},
distillation-based methods \cite{salimans2022progressivedistillationfastsampling},
and consistency training \cite{song2023consistencymodels,luo2023latent}.
These methods improve latency primarily by shortening
the sampling trajectory itself,
often at the cost of additional training,
specialized solvers,
or re-parameterized objectives.

\textbf{Diffusion Quantization.}
Quantization compresses diffusion transformers
to reduce memory footprint and latency.
Post-training quantization (PTQ) methods
optimize deployment without retraining,
including PTQ4DiT \cite{wu2024ptq4dit},
which addresses outliers and timestep-wise distribution shifts,
enabling training-free low-bit quantization.
SVDQuant \cite{li2024svdquant}
uses low-rank branches to absorb outliers,
achieving 4-bit quantization with minimal quality loss.
For video models,
Q-VDiT \cite{feng2025q}
employs token-aware quantization and temporal distillation,
while ViDiT-Q \cite{zhao2024viditq}
adopts fine-grained dynamic quantization and mixed precision strategies.

\textbf{Feature Reuse.}
Given the computational overhead of training-dependent acceleration,
recent work has focused on training-free methods
that generalize across diverse diffusion models.
Feature reuse exploits activation redundancy across denoising steps
through intermediate caching and reuse,
offering superior generalizability.

Early approaches apply caching to U-Net-based models:
DeepCache \cite{Xu_2018},
Cache Me if You Can \cite{wimbauer2024cachecanacceleratingdiffusion},
Faster Diffusion \cite{li2024faster},
FasterCache \cite{lv2024fastercache},
and T-GATE \cite{tgate}
explore selective reuse at the block, branch, or attention level,
showing that substantial redundancy exists
across denoising iterations in image and video diffusion models.
Extending this idea to Diffusion Transformers,
FORA \cite{selvaraju2024forafastforwardcachingdiffusion},
ToCa \cite{zou2024accelerating},
DuCa \cite{zou2025rethinkingtokenwisefeaturecaching},
$\Delta$-DiT \cite{chen2024deltadittrainingfreeaccelerationmethod},
and AdaCache \cite{adacache}
largely rely on heuristic or token-wise caching policies
to reuse attention and MLP activations.

Prediction-based reuse methods take a different route.
TeaCache \cite{liu2024timestep}
fits model-specific polynomial coefficients from calibration data,
whereas MagCache \cite{ma2025magcache}
uses precomputed magnitude-ratio schedules
to guide reuse decisions.
TaylorSeer \cite{TaylorSeer2025},
HiCache \cite{feng2025hicachetrainingfreeaccelerationdiffusion},
and AB-Cache \cite{yu2025abcachetrainingfreeaccelerationdiffusion}
forecast future activations
through fixed-form extrapolation schemes,
including expansions based on Taylor, Hermite, and Adams--Bashforth formulations.
These methods further improve reuse quality,
but their forecasting behavior remains tied
to preselected analytic forms or offline-calibrated control profiles.

\textbf{Distinction from Prior Work:}
$\mathrm{D}^3$-Cache fundamentally differs from existing feature reuse methods
through its foundation in Dynamic Mode Decomposition.
Rather than relying on model-specific,
offline-calibrated control profiles
or fixed-form extrapolation schemes,
$\mathrm{D}^3$-Cache estimates a local evolution operator online
from recent transformer outputs.
This yields a data-driven forecasting mechanism
that adapts to local feature dynamics
and supports error-aware skipping of redundant computations,
leading to a stronger quality-efficiency trade-off
than prior cache-based acceleration methods.
